---
title: "dieselPM_TSSTcort_CDI"
author: "Jessica Uy and Katy Shin"
date: '2024-05-20'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## load libraries
```{r}
library(dplyr)
library(readxl)
library(tidyverse)
library(psych)
library(quest)
library(lme4)
library(naniar)
library(corrplot)
library(ggplot2)
library(sjPlot)
library(car)
```

## load files
```{r}
dieselcort <- read_csv("dieselPM_TSSTcort_pub.csv")
cort_compare <- read_csv("ELS_cort_compare_pub.csv")
```

## winsorize cort data exceeding 2SD
```{r}
dieselcort_win <-
  dieselcort %>%
  mutate(CVS1_winsorized_2sd = quest::winsor(dieselcort$CVS1, z.max = 2, rtn.int = FALSE, to.na = FALSE)) %>%
  mutate(CVS2_winsorized_2sd = quest::winsor(dieselcort$CVS2, z.max = 2, rtn.int = FALSE, to.na = FALSE)) %>%
  mutate(CVS3_winsorized_2sd = quest::winsor(dieselcort$CVS3, z.max = 2, rtn.int = FALSE, to.na = FALSE)) %>%
  mutate(CVS4_winsorized_2sd = quest::winsor(dieselcort$CVS4, z.max = 2, rtn.int = FALSE, to.na = FALSE)) %>%
  mutate(CVS5_winsorized_2sd = quest::winsor(dieselcort$CVS5, z.max = 2, rtn.int = FALSE, to.na = FALSE))
  
```

## calculate AUCg
based on formula from Pruessner et al. (2003)
```{r}
dieselcort_win_AUC <-
  dieselcort_win %>%
  # AUCg_2to4 = cortisol reactivity
   mutate(AUCg_2to4 = ((CVS2_winsorized_2sd + CVS3_winsorized_2sd)*TDM.S2.S3)/2 + ((CVS3_winsorized_2sd + CVS4_winsorized_2sd)*TDM.S3.S4)/2) %>%
  # AUCg_4to5 = cortisol recovery
   mutate(AUCg_4to5 = ((CVS4_winsorized_2sd + CVS5_winsorized_2sd)*TDM.S4.S5)/2)
```

## calculate composites
- neighborhood socioeconomic status indicators were z-scored and averaged to create the neighborhood socioeconomic disadvantage composite
- parent education and household income were z-scored and averaged to create a household SES composite and multiplied by -1
- diesel PM percentiles from CalEnviroScreen (CES) 3.0 and 4.0 were z-scored and averaged
- PM2.5 percentiles from CES 3.0 and 4.0 were z-scored and averaged

```{r}
dieselcort_win_AUC_SES <-
   dieselcort_win_AUC %>%
  mutate(poverty_pctl_zscore = (poverty_pctl - mean(poverty_pctl,na.rm=T))/sd(poverty_pctl, na.rm=T)) %>%
  mutate(unemployment_pctl_zscore = (unemployment_pctl - mean(unemployment_pctl, na.rm=T))/sd(unemployment_pctl, na.rm=T)) %>%
  mutate(housing_burden_pctl_zscore = (housing_burden_pctl - mean(housing_burden_pctl, na.rm=T))/sd(housing_burden_pctl, na.rm=T)) %>%
  mutate(education_pctl_zscore = (education_pctl - mean(education_pctl, na.rm=T))/sd(education_pctl, na.rm=T)) %>%
  mutate(neighborhood_SES = (poverty_pctl_zscore + unemployment_pctl_zscore + housing_burden_pctl_zscore + education_pctl_zscore)/4) %>%
  mutate(parent_education_zscore = (Parent_Education - mean(Parent_Education,na.rm=T))/sd(Parent_Education, na.rm=T)) %>%
  mutate(parent_income_zscore = (Parent_Income - mean(Parent_Income, na.rm=T))/sd(Parent_Income, na.rm=T)) %>%
  mutate(family_SES = (parent_education_zscore + parent_income_zscore)/2) %>%
  mutate(family_SES_reverse = -1*family_SES) %>%
  mutate(diesel_ces3_zscore = (diesel_PM_pctl - mean(diesel_PM_pctl, na.rm=T))/sd(diesel_PM_pctl, na.rm=T)) %>%
  mutate(diesel_ces4_zscore = (diesel_PM_pctl_4 - mean(diesel_PM_pctl_4, na.rm=T))/sd(diesel_PM_pctl_4, na.rm=T)) %>%
  mutate(diesel_PM_pctl_z = (diesel_ces3_zscore + diesel_ces4_zscore)/2) %>% 
  mutate(PM2.5_pctl_zscore = (PM2.5_pctl - mean(PM2.5_pctl, na.rm=T))/sd(PM2.5_pctl, na.rm=T)) %>%
  mutate(PM2.5_pctl_zscore_4 = (PM2.5_pctl_4 - mean(PM2.5_pctl_4, na.rm=T))/sd(PM2.5_pctl_4, na.rm=T)) %>%
  mutate(PM2.5_pctlavg_z = (PM2.5_pctl_zscore + PM2.5_pctl_zscore_4)/2)
```

## correlation table
```{r}
dieselcort_win_AUC_SES_corr <-
  dieselcort_win_AUC_SES %>%
  select(Age, Tanner_Score,family_SES_reverse, neighborhood_SES, diesel_PM_pctl_z, AUCg_2to4, AUCg_4to5, CDI_total, PM2.5_pctlavg_z, MASC_total, YSR_Ext_Problems_Total)

corPlot(dieselcort_win_AUC_SES_corr,
        numbers=TRUE,
        main = "",
        stars=TRUE,
        show.legend=TRUE,
        diag=FALSE,
        upper=FALSE,
        xlas=2,
        cex.axis = 0.5,
        cex = 0.5)
```

## linear regressions
```{r}
diesel_CDI <- lm(CDI_total ~ diesel_PM_pctl_z + Age + Sex + Tanner_Score + batch + family_SES_reverse + neighborhood_SES, data=dieselcort_win_AUC_SES)
summary(diesel_CDI)
tab_model(diesel_CDI)

# convert t values to Cohen's d values for CDI_total ~ diesel_PM_pctl_z
t_to_d(t = 2.004, df_error = 150)

diesel_AUCg2to4 <- lm(AUCg_2to4 ~ diesel_PM_pctl_z + Age + Sex + Tanner_Score + batch + family_SES_reverse + neighborhood_SES, data=dieselcort_win_AUC_SES)
summary(diesel_AUCg2to4)
tab_model(diesel_AUCg2to4)

# convert t values to Cohen's d values for AUCg_2to4 (reactivity) ~ diesel_PM_pctl_z
t_to_d(t = -2.406, df_error = 147)

diesel_AUCg4to5 <- lm(AUCg_4to5 ~ diesel_PM_pctl_z + Age + Sex + Tanner_Score + batch + family_SES_reverse + neighborhood_SES, data=dieselcort_win_AUC_SES)
summary(diesel_AUCg4to5)
tab_model(diesel_AUCg4to5)

# convert t values to Cohen's d values for AUCg_4to5 (recovery) ~ diesel_PM_pctl_z
t_to_d(t = -1.805, df_error = 149)

AUCg2to4_CDI <- lm(CDI_total ~ AUCg_2to4 + diesel_PM_pctl_z + Age + Sex + Tanner_Score + batch + family_SES_reverse + neighborhood_SES, data=dieselcort_win_AUC_SES)
summary(AUCg2to4_CDI)
tab_model(AUCg2to4_CDI)

# convert t values to Cohen's d values for CDI_total ~ AUCg reactivity
t_to_d(t = -2.05, df_error = 146)

AUCg4to5_CDI <- lm(CDI_total ~ AUCg_4to5 + diesel_PM_pctl_z + Age + Sex + Tanner_Score + batch + family_SES_reverse + neighborhood_SES, data=dieselcort_win_AUC_SES)
summary(AUCg4to5_CDI)
tab_model(AUCg4to5_CDI)
```

## mediation analysis: diesel PM --> AUCg Reactivity --> CDI total
run process.R to use process function
```{r}
process(data=dieselcort_win_AUC_SES, y="CDI_total", x="diesel_PM_pctl_z", m=c("AUCg_2to4"), cov=c("Age", "Sex", "neighborhood_SES", "family_SES_reverse", "Tanner_Score", "batch"), model=4, seed=100770, conf=95, intprobe=1)
```

## set up plot theme
```{r}
set_theme(
  geom.outline.color = "black", 
  geom.outline.size = 1, 
  geom.label.size = 2,
  geom.label.color = "black",
  title.color = "black", 
  title.size = 1.5, 
  axis.angle.x = 0, 
  axis.textcolor = "black", 
  base = theme_bw()
)
```

## create plots for mediation scatterplots
```{r}
plot_model(diesel_AUCg2to4,
           type = "pred",
           terms = c("diesel_PM_pctl_z"),
           colors = c("black", "black"),
           show.data = TRUE,
           dot.size = 3,
          jitter = 0.3,
           show.p = TRUE) +
    labs(title = "", 
        x = "Diesel PM Percentile (Standardized Average)", 
        y = "Cortisol Reactivity (AUCg)")

plot_model(AUCg2to4_CDI,
           type = "pred",
           terms = c("AUCg_2to4"),
           colors = c("black", "black"),
           show.data = TRUE,
           dot.size = 3,
          jitter = 0.3,
           show.p = TRUE) +
    labs(title = "", 
        x = "Cortisol Reactivity (AUCg)", 
        y = "Depressive Symptoms")
```

## supplemental analyses cort samples
```{r}
diesel_S2 <- lm(CVS2_winsorized_2sd ~ diesel_PM_pctl_z + Age + Sex + Tanner_Score + batch + family_SES_reverse + neighborhood_SES, data=dieselcort_win_AUC_SES)
summary(diesel_S2)
tab_model(diesel_S2)
t_to_d(t = -1.765, df_error = 151)

diesel_S3 <- lm(CVS3_winsorized_2sd ~ diesel_PM_pctl_z + Age + Sex + Tanner_Score + batch + family_SES_reverse + neighborhood_SES, data=dieselcort_win_AUC_SES)
summary(diesel_S3)
tab_model(diesel_S3)
t_to_d(t = -2.263, df_error = 151)

diesel_S4 <- lm(CVS4_winsorized_2sd ~ diesel_PM_pctl_z + Age + Sex + Tanner_Score + batch + family_SES_reverse + neighborhood_SES, data=dieselcort_win_AUC_SES)
summary(diesel_S4)
tab_model(diesel_S4)
t_to_d(t = -1.243, df_error = 150)

diesel_S5 <- lm(CVS5_winsorized_2sd ~ diesel_PM_pctl_z + Age + Sex + Tanner_Score + batch + family_SES_reverse + neighborhood_SES, data=dieselcort_win_AUC_SES)
summary(diesel_S5)
tab_model(diesel_S5)
t_to_d(t = -1.458, df_error = 150)

```

## calculate predicted cort values per sample at high (+1SD) and low (-1SD) of diesel PM percentile
graph of predicted cort values and SE for each saliva sample at high and low levels of diesel PM percentile were plotted using Excel
```{r}
newdata <- data.frame(diesel_PM_pctl_z=c(-0.9, 0, 0.90), # 1SD = 0.9
Age=c(12.26112, 12.26112, 12.26112), # mean age
Sex=c(1.564706, 1.564706, 1.564706), # mean sex
Tanner_Score=c(2.7, 2.7, 2.7), # mean Tanner
neighborhood_SES=c(-0.00854691, -0.00854691, -0.00854691), # mean neighborhood SES
batch=c(0.5352941, 0.5352941, 0.5352941), # mean batch
family_SES_reverse=c(0, 0, 0)) # mean household SES

predict(diesel_S2, newdata, se.fit=TRUE) # obtain predicted cort values and SE for S2
predict(diesel_S3, newdata, se.fit=TRUE) # obtain predicted cort values and SE for S3
predict(diesel_S4, newdata, se.fit=TRUE) # obtain predicted cort values and SE for S4
predict(diesel_S5, newdata, se.fit=TRUE) # obtain predicted cort values and SE for S5
```

## specificity analyses
```{r}
## diesel PM --> CDI --> AUCg reactivity
process(data=dieselcort_win_AUC_SES, y="AUCg_2to4", x="diesel_PM_pctl_z", m=c("CDI_total"), cov=c("Age", "Sex", "neighborhood_SES", "family_SES_reverse", "Tanner_Score", "batch"), model=4, seed=100770, conf=95, intprobe=1)

## diesel PM --> AUCg reactivity --> MASC
process(data=dieselcort_win_AUC_SES, y="MASC_total", x="diesel_PM_pctl_z", m=c("AUCg_2to4"), cov=c("Age", "Sex", "neighborhood_SES", "family_SES_reverse", "Tanner_Score", "batch"), model=4, seed=100770, conf=95, intprobe=1)

## diesel PM --> AUCg reactivity --> YSR externalizing
process(data=dieselcort_win_AUC_SES, y="YSR_Ext_Problems_Total", x="diesel_PM_pctl_z", m=c("AUCg_2to4"), cov=c("Age", "Sex", "neighborhood_SES", "family_SES_reverse", "Tanner_Score", "batch"), model=4, seed=100770, conf=95, intprobe=1)

process(data=dieselcort_win_AUC_SES, y="CDI_total", x="PM2.5_pctlavg_z", m=c("AUCg_2to4"), cov=c("Age", "Sex", "neighborhood_SES", "family_SES_reverse", "Tanner_Score", "batch"), model=4, seed=100770, conf=95, intprobe=1)
```

## comparison of baseline demographic data between individuals who did not do TSST vs. those who did
```{r}
nocort <-
  cort_compare %>%
  filter(has_cort==0)

yescort <-
  cort_compare %>%
  filter(has_cort==1)

# get descriptives
nocort_desc <- psych::describe(nocort)
yescort_desc <- psych:: describe(yescort)

# t-tests
t.test(Tanner_Average_T1 ~ has_cort, data = cort_compare)

t.test(Age_T1 ~ has_cort, data = cort_compare)

t.test(Parent_Education_T1 ~ has_cort, data = cort_compare)

t.test(Parent_Income_T1 ~ has_cort, data = cort_compare)

t.test(CDI_total_T1 ~ has_cort, data=cort_compare)

t.test(PM2.5_pctl ~ has_cort, data=cort_compare)

t.test(diesel_PM_pctl ~ has_cort, data=cort_compare)

t.test(education_pctl ~ has_cort, data=cort_compare)

t.test(poverty_pctl ~ has_cort, data=cort_compare)

t.test(unemployment_pctl ~ has_cort, data=cort_compare)

t.test(housing_burden_pctl ~ has_cort, data=cort_compare)

chisq.test(cort_compare$Sex, cort_compare$has_cort)

chisq.test(cort_compare$Race, cort_compare$has_cort)

```

